---
title: "Portfolio 4 - Openness and Well‑Being in MTurk Adults: Personality Correlates Informing an AI Chatbot Future‑Clarity Study"
---

### Goal 

My follow up study is adapting from a classmate's study done on openness to experience, which found that individuals had a greater sense of their purpose in life after talking to an AI-chatbot on purpose discovery (they had a higher purpose in life measure), and this effect was greater in individuals with higher openness to experience. Thus, this portfolio piece will highlight as building background theory for the chatbot work. Together with personality data is support of the general idea: openness makes people more willing to explore, tolerate ambiguity, and try out new identity narratives in a semi‑novel context (an AI conversation). That mechanism is naturally extendable from “sense of purpose” to “future‑plan clarity,” because both involve imagining, evaluating, and committing to future directions. My hope is that I can then reasonably claim if openness amplifies gains in purpose after chatbot reflection, and openness is tied to self‑esteem but not just to general life satisfaction or demographics, then openness should also amplify gains in future‑plan clarity from a similar chatbot conversation.

The primary goal of this portfolio piece is to visualize how openness to experience relates to self‑esteem, life satisfaction, need for cognition, age, and gender in an MTurk adult sample, and use this “openness profile” to motivate why openness should moderate gains in future‑plan clarity after an AI purpose‑discovery chatbot conversation.


### Products

For this portfolio, I will create several products that together tell a clear story about openness in an MTurk adult sample. First, I will generate a descriptive statistics table summarizing openness, self‑esteem, satisfaction with life, need for cognition, age, and gender proportions. Next, I will present a polished correlation matrix figure that highlights how openness relates positively to self‑esteem and need for cognition, but is largely independent of life satisfaction, age, and gender. I will then include a set of scatterplots with regression lines (openness with self‑esteem, need for cognition, and satisfaction) to visually contrast meaningful versus near‑zero relationships. In addition, I will show a gender comparison plot (boxplot or violin) for openness, and optionally self‑esteem and need for cognition, to illustrate the minimal gender differences in these traits.


```{r - loading data}
library(tidyverse)
library(haven)

mTurk_raw <- read_sav("OnlinePsychologySurveyData.sav") %>% 
  as_tibble()
```

```{r - create dataframe to measure openness, self-esteem, satisfaction, nfc, age, and gender scores}
mTurk_correlates <- mTurk_raw %>% 
  transmute(
    openness           = OPENNESS,
    self_esteem    = SELFESTEEM,         
    satisfaction   = SATISFACTION,
    nfc = NFC,
    age = Age,
    gender = Gender
  )
```

```{r - recoding gender}
mTurk_correlates <- mTurk_correlates %>% 
  mutate(
    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )

glimpse(mTurk_correlates)
```
```{r - descriptive stats}
mTurk_correlates %>% 
  summarise(
    n           = n(),
    mean_open   = mean(openness, na.rm = TRUE),
    sd_open     = sd(openness, na.rm = TRUE),
    mean_se     = mean(self_esteem, na.rm = TRUE),
    sd_se       = sd(self_esteem, na.rm = TRUE),
    mean_satis  = mean(satisfaction, na.rm = TRUE),
    sd_satis    = sd(satisfaction, na.rm = TRUE),
    mean_nfc    = mean(nfc, na.rm = TRUE),
    sd_nfc      = sd(nfc, na.rm = TRUE),
    mean_age    = mean(age, na.rm = TRUE),
    sd_age      = sd(age, na.rm = TRUE)
  )

mTurk_correlates %>% 
  count(gender)
```

```{r - correlation matrix}
cor_vars <- mTurk_correlates %>% 
  select(openness, self_esteem, satisfaction, nfc, age)

cor_matrix <- cor(cor_vars, use = "pairwise.complete.obs")
cor_matrix
```

```{r - loading psych}
library(psych)

psych::corr.test(cor_vars)

cor_results <- psych::corr.test(cor_vars)
cor_results$r      # correlation coefficients
cor_results$p      # p-values

```

```{r}
library(ggplot2)
```

```{r - openness x self esteem}
ggplot(mTurk_correlates, aes(x = openness, y = self_esteem)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "brown") +
  labs(
    x = "Openness to Experience",
    y = "Self-esteem",
    title = "Openness and Self-esteem in MTurk Adults"
  )
```

Interpretation: Higher openness scores are associated with higher self‑esteem in this MTurk adult sample. As openness increases from about 2 to 7, the regression line shows a steady upward slope, indicating that more open participants tend to feel better about themselves overall. This is supportive of the theory that openness is linked to more self-confidence, which can support in explaining why individuals with higher openness might engage in a more constructive manner with an AI chatbot about their future plans. 


```{r - openness x satisfaction}
ggplot(mTurk_correlates, aes(x = openness, y = satisfaction)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "brown") +
  labs(
    x = "Openness to Experience",
    y = "Satisfaction",
    title = "Openness and Satisfaction with Life in MTurk Adults"
  )
```


Interpretation: This plot shows that openness is essentially unrelated to life satisfaction in your MTurk sample. The regression line is almost flat, and points are widely scattered across satisfaction levels at every openness score. More open adults do not report being generally more satisfied with their lives.


```{r - openness x NFC}  
ggplot(mTurk_correlates, aes(x = openness, y = nfc)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "brown") +
  labs(
    x = "Openness to Experience",
    y = "Need for Cognition",
    title = "Openness and Need for Cognition in MTurk Adults"
  )
```


Interpretation: This figure shows a strong positive association between openness and need for cognition in your MTurk adults. As openness increases, the regression line rises steeply, and higher‑openness participants tend to report clearly higher enjoyment of effortful thinking. More open adults are not just curious; they also like engaging with complex, cognitively demanding tasks.



```{r - gender and openness}
ggplot(mTurk_correlates, aes(x = gender, y = openness, fill = gender)) +
  geom_boxplot() +
  scale_fill_manual(values = c("Male" = "red",
                               "Female" = "green")) +
  labs(
    x = "Gender",
    y = "Openness to Experience",
    title = "Openness by Gender in MTurk Adults"
  )
```


Interpretation: Openness to experience is very similar for male and female MTurk workers in the sample. The medians for both genders are around the same value, with overlapping interquartile ranges and whiskers, indicating no substantial gender difference in openness scores. Openness appears to cut across gender in this adult online sample, and this reinforces the idea that it’s a psychological trait rather than a demographic marker.



```{r - facet of openness x correlates by gender}
mTurk_long <- mTurk_correlates %>% 
  filter(!is.na(gender)) %>% 
  pivot_longer(
    cols = c(self_esteem, satisfaction, nfc),
    names_to = "outcome",
    values_to = "score"
  ) %>% 
  mutate(
    outcome = factor(
      outcome,
      levels = c("self_esteem", "satisfaction", "nfc"),
      labels = c("Self-esteem", "Satisfaction with life", "Need for cognition")
    )
  )

ggplot(mTurk_long,
       aes(x = gender, y = score, fill = gender)) +
  geom_boxplot(alpha = 0.6, width = 0.6) +
  facet_wrap(~ outcome, nrow = 1) +
  scale_fill_manual(values = c("Male" = "red", "Female" = "green")) +
  labs(
    x = "Gender",
    y = "Scale score",
    title = "Self-esteem, Satisfaction, and Need for Cognition by Gender in MTurk Adults"
  ) +
  theme(legend.position = "none") +
  theme_minimal()
```

Interpretation: Self‑esteem, life satisfaction, and need for cognition are broadly similar for male and female individuals In all three panels, the male and female boxplots have overlapping medians and interquartile ranges, with only small differences at most. Both openness and its key correlates (self‑esteem, satisfaction, NFC) are not strongly gendered in this sample.


### Data

The dataset comes from an online survey of 300 adult workers recruited through Amazon Mechanical Turk, who completed measures of openness to experience, self‑esteem, satisfaction with life, need for cognition, and basic demographics. This data was distributed in my Research Methods 1 class. Openness, self‑esteem, satisfaction, and NFC were each scored as averaged Likert‑type scales, with higher scores indicating greater levels of the underlying trait or attitude. Age was recorded in years (continuous), and gender was self‑reported using a binary male/female option in this sample.


### Interpretation


Openness to experience in this MTurk adult sample is positively related to both self‑esteem and need for cognition, but essentially unrelated to global life satisfaction, age, or gender. Adults who score higher on openness tend to feel somewhat better about themselves and enjoy effortful thinking more, yet they are not simply happier overall, older, or drawn from one gender group. This pattern supports treating openness (and related cognitive engagement) as a distinct psychological resource that cuts across demographics, which is why it is predicted that more open individuals will show larger gains in future‑plan clarity after engaging with a purpose‑discovery AI chatbot.




