---
title: "Portfolio 1"
---

  > The projects should be numbered consecutively (i.e., in the order in which you began them), and should include for each project a description of the goal, the product (computer program, hand graph, computer graph, etc.), the data, and some interpretation. Reports must be reproducible and of high quality in terms of writing, grammar, presentation, etc.

Goal: In this project, I explore how often leading AI chatbots repeat false claims when asked about news-related misinformation, compared to how often they debunk or refuse to answer. I chose this because my portfolio focuses on AI as a source for current events and how that shapes exposure to misinformation.

Product: The main product is a set of ggplot2 graphs: (1) a bar chart showing the distribution of chatbot responses (false claim vs debunk vs non-response) in a recent audit; and (2) an extended version that includes an earlier audit so I can visualize changes over time.

Data: 
```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidyr)
```

```{r}
chatbot_audits <- read_csv("chatbot_audits.csv")
```
I constructed a small tidy dataset from NewsGuard’s AI Misinformation Monitor and related summaries of disinformation rates in leading chatbots. For each audit period, I recorded the percentage of prompts that led to false claims, non-responses, or debunks/corrections. I entered these values into a CSV and imported them into R.

Interpretation:



```{r}
# Create a barplot of the number of fish encounters by species
ggplot(data = fish_encounters, aes(x = station, y = seen)) +
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Number of Fish Encounters")
  
```


Sed mollis purus faucibus leo finibus, ut pharetra nibh elementum. Nulla id orci lorem. Mauris sollicitudin mi eget tellus vehicula, ut convallis lorem mollis. In blandit nulla diam, a malesuada sem volutpat interdum. Ut elementum elit at velit maximus, ac tempor odio porttitor. Suspendisse lobortis sollicitudin tortor, ut pharetra libero luctus a. Sed hendrerit faucibus metus. Sed rutrum purus vel pellentesque euismod.


```{r}
# Create a barplot of the number of fish encounters by species
ggplot(data = fish_encounters, aes(x = station, y = seen)) +
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Number of Fish Encounters")
  
```

Mauris placerat in velit a pharetra. Nulla viverra bibendum lectus, ut lobortis nibh. Aenean malesuada, orci vitae dictum placerat, enim justo aliquet leo, ac tincidunt augue ipsum et mi. Nullam in pharetra metus, nec molestie felis. Duis varius rhoncus pulvinar. Integer semper urna sed tellus rhoncus, vel suscipit odio iaculis. Mauris iaculis neque pellentesque ex convallis venenatis. Sed ultrices euismod fermentum. Ut feugiat risus at turpis tincidunt viverra.


```{r}
# Create a barplot of the number of fish encounters by species
ggplot(data = fish_encounters, aes(x = station, y = seen)) +
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Number of Fish Encounters")
  
```

Donec sagittis suscipit orci in ultrices. Vivamus iaculis eleifend est et dictum. Fusce non urna at lacus dapibus tincidunt ac sed nibh. Donec luctus lectus elit, ac ultrices tortor mollis sit amet. Nulla finibus lacinia arcu. Proin sollicitudin non odio et pulvinar. Maecenas finibus faucibus finibus. Ut elementum metus velit, a sodales sapien commodo et. Vivamus pulvinar, sapien et porta cursus, elit est luctus nisi, eget fringilla odio diam et enim. Sed sodales suscipit bibendum. Morbi ut metus semper, laoreet felis sodales, consequat dolor.
