---
title: "Portfolio 3 - Individual Differences Within AI Question Framing and Emotional
  Responses to Purpose Reflection in College Upperclassmen"
---

### Goal


In this project, I build off of my earlier portfolio piece to explore how negative affect varies across individual‑difference variables in my AI post‑graduation planning study, focusing on gender, class year, and age as a baseline descriptive step before I add psychological traits like openness in later work. With this, I want to practice building a clean analytic tibble from my SPSS export, create multiple ggplot visualizations of distress across groups, and add simple numeric summaries that connect directly to the figures.  The original study recruited 60 U.S. juniors and seniors to report their post‑graduation plan clarity, respond to either two specific or two open‑ended AI‑generated purpose questions, and then rate their negative affect, with the main finding that higher plan clarity predicted lower distress while AI question type and its interaction with clarity were non‑significant. 



### Products

My products will include a cleaned analytic tibble that contains negative affect, gender, class year, and age, with gender and class year recoded into factor labels that are readable and removing any rows with values missing so the same complete‑case sample is used for all of the analyses. With this tibble, I created three visualizations: a plot of negative affect by gender, a plot of negative affect by class year, and a scatterplot of age versus negative affect with a smooth line to show any trend in distress across the narrow upper‑division age range. In addition to the graphs, I generated small summary tibbles reporting the mean and standard deviation of negative affect by gender and by class year, plus the overall mean, standard deviation, and range of negative affect for the full sample, so I can move between visual and numeric descriptions of the same patterns.

```{r - loading data}
library(tidyverse)
library(haven)

studydata_raw <- read_sav("AI Purpose Questions and Plan Clarity Study.sav") %>% 
  as_tibble()
```

```{r - create dataframe to measure emotional distress scores}
studydata_emotion <- studydata_raw %>% 
  transmute(
    id           = ID,
    condition    = condition,         
    neg_affect   = NegAff_Mean
  )
```

```{r - creating new demographics dataframe}
gender_demo <- studydata_raw %>% 
  transmute(
    id     = ID,
    gender = Gender
  )

classyear_demo <- studydata_raw %>% 
  transmute(
    id         = ID,
    class_year = Study_Year
  )

age_demo <- studydata_raw %>% 
  transmute(
    id  = ID,
    age = Age
  )
```

```{r - joining demographics dataframes to neg affect}
studydata_emotion_demo <- studydata_emotion %>% 
  left_join(gender_demo,    by = "id") %>% 
  left_join(classyear_demo, by = "id") %>% 
  left_join(age_demo,       by = "id")
```

```{r - numeric summary for overall distress}
overall_distress_summary <- studydata_emotion_demo %>% 
  summarise(
    n        = n(),
    mean     = mean(neg_affect),
    sd       = sd(neg_affect),
    min      = min(neg_affect),
    max      = max(neg_affect)
  )
```

```{r - numeric summary for gender}
gender_distress_summary <- studydata_emotion_demo %>% 
  group_by(gender) %>% 
  summarise(
    n    = n(),
    mean = mean(neg_affect),
    sd   = sd(neg_affect),
    .groups = "drop"
  )
```

```{r - numeric summary for class year}
classyear_distress_summary <- studydata_emotion_demo %>% 
  group_by(class_year) %>% 
  summarise(
    n    = n(),
    mean = mean(neg_affect),
    sd   = sd(neg_affect),
    .groups = "drop"
  )
```

```{r - merging numerical summaries}
combined_distress_summary <- bind_rows(
  overall_distress_summary %>% 
    mutate(
      grouping = "overall",
      level    = "all"
    ),
  gender_distress_summary %>% 
    mutate(
      grouping = "gender",
      level    = as.character(gender)
    ) %>% 
    select(grouping, level, n, mean, sd),
  classyear_distress_summary %>% 
    mutate(
      grouping = "class_year",
      level    = as.character(class_year)
    ) %>% 
    select(grouping, level, n, mean, sd)
) %>% 
  relocate(grouping, level)
```

```{r - distress by gender plot}
p_distress_gender <- ggplot(studydata_emotion_demo,
                            aes(x = gender, y = neg_affect)) +
  geom_boxplot(fill = "steel blue", alpha = 0.7) +
  labs(
    title = "Negative Affect by Gender",
    x     = "Gender",
    y     = "Negative affect (1–5)"
  ) +
  theme_minimal()

p_distress_gender
```

Interpretation: It seems like the median distress score for females are slightly higher than males, with greater variability among women. The shorter boxplot of the males suggest that most men seem to cluster around a relative low affect. The two boxplots share a strong overlap, implying that gender is not a strong implicator in emotional distress for this sample.

```{r - distress by class year plot}
p_distress_classyear <- ggplot(studydata_emotion_demo,
                               aes(x = class_year,
                                   y = neg_affect,
                                   fill = factor(class_year))) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(
    name   = "Class year",
    values = c("3" = "steel blue", "4" = "light green"),
    labels = c("3" = "Junior", "4" = "Senior")
  ) +
  labs(
    title = "Negative Affect by Class Year",
    x     = "Class year",
    y     = "Negative affect (1–5)"
  ) +
  theme_minimal()

p_distress_classyear
```


Interpretation: Juniors and seniors seem to report similar overall levels of negative affect. The median for seniors are slightly higher, which may possibly be from feeling a bit more distress on average. I would say that the variation between the two class years is larger than differences between them. 


