---
title: "Portfolio 4 - Openness and Well‑Being in MTurk Adults: Personality Correlates Informing an AI Chatbot Future‑Clarity Study"
---

### Goal 

My follow up study is adapting from a classmate's study done on openness to experience, which found that individuals had a greater sense of their purpose in life after talking to an AI-chatbot on purpose discovery (they had a higher purpose in life measure), and this effect was greater in individuals with higher openness to experience. Thus, this portfolio piece will highlight as building background theory for the chatbot work. Together with personality data is support of the general idea: openness makes people more willing to explore, tolerate ambiguity, and try out new identity narratives in a semi‑novel context (an AI conversation). That mechanism is naturally extendable from “sense of purpose” to “future‑plan clarity,” because both involve imagining, evaluating, and committing to future directions. My hope is that I can then reasonably claim if openness amplifies gains in purpose after chatbot reflection, and openness is tied to self‑esteem but not just to general life satisfaction or demographics, then openness should also amplify gains in future‑plan clarity from a similar chatbot conversation.

The primary goal of this portfolio piece is to visualize how openness to experience relates to self‑esteem, life satisfaction, need for cognition, age, and gender in an MTurk adult sample, and use this “openness profile” to motivate why openness should moderate gains in future‑plan clarity after an AI purpose‑discovery chatbot conversation.


### Products

For this portfolio, I will create several products that together tell a clear story about openness in an MTurk adult sample. First, I will generate a descriptive statistics table summarizing openness, self‑esteem, satisfaction with life, need for cognition, age, and gender proportions. Next, I will present a polished correlation matrix figure that highlights how openness relates positively to self‑esteem and need for cognition, but is largely independent of life satisfaction, age, and gender. I will then include a set of scatterplots with regression lines (openness with self‑esteem, need for cognition, and satisfaction) to visually contrast meaningful versus near‑zero relationships. In addition, I will show a gender comparison plot (boxplot or violin) for openness, and optionally self‑esteem and need for cognition, to illustrate the minimal gender differences in these traits.


```{r - loading data}
library(tidyverse)
library(haven)

mTurk_raw <- read_sav("OnlinePsychologySurveyData.sav") %>% 
  as_tibble()
```

```{r - create dataframe to measure openness, self-esteem, satisfaction, nfc, age, and gender scores}
mTurk_correlates <- mTurk_raw %>% 
  transmute(
    openness           = OPENNESS,
    self_esteem    = SELFESTEEM,         
    satisfaction   = SATISFACTION,
    nfc = NFC,
    age = Age,
    gender = Gender
  )
```

```{r - recoding gender}
mTurk_correlates <- mTurk_correlates %>% 
  mutate(
    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )

glimpse(mTurk_correlates)
```
```{r - descriptive stats}
mTurk_correlates %>% 
  summarise(
    n           = n(),
    mean_open   = mean(openness, na.rm = TRUE),
    sd_open     = sd(openness, na.rm = TRUE),
    mean_se     = mean(self_esteem, na.rm = TRUE),
    sd_se       = sd(self_esteem, na.rm = TRUE),
    mean_satis  = mean(satisfaction, na.rm = TRUE),
    sd_satis    = sd(satisfaction, na.rm = TRUE),
    mean_nfc    = mean(nfc, na.rm = TRUE),
    sd_nfc      = sd(nfc, na.rm = TRUE),
    mean_age    = mean(age, na.rm = TRUE),
    sd_age      = sd(age, na.rm = TRUE)
  )

mTurk_correlates %>% 
  count(gender)
```

```{r - correlation matrix}
cor_vars <- mTurk_correlates %>% 
  select(openness, self_esteem, satisfaction, nfc, age)

cor_matrix <- cor(cor_vars, use = "pairwise.complete.obs")
cor_matrix
```






