---
title: "Portfolio 5 - Openness and the Big Five in Emerging Adulthood"
---

### Goal

Expand on my previous portfolio piece looking at a sample from mTurk to observe patterns to further my future work on openness to experience. This portfolio piece will focus to test how well other Big Five traits (extraversion, agreeableness, conscientiousness, and emotional stability) predict openness to experience, and examine whether mean levels and predictors of openness are similar across emerging‑adult age groups (18–23, 23–25, 25–30) that parallel the target ages in my planned chatbot study.


### Products

This project will produce a cleaned Big Five dataset with TIPI‑based composite scores for extraversion, agreeableness, conscientiousness, emotional stability, and openness. I will create a correlation table showing how openness relates to the other four traits in the full MTurk sample, followed by a multiple regression model predicting openness from extraversion, agreeableness, conscientiousness, and emotional stability with standardized coefficients and overall model fit. I will derive age variables that capture the emerging‑adult bands 18–23, 23–25, and 25–30, and use these to compute and visualize mean openness and confidence intervals across age bands. Finally, I will run parallel regressions predicting openness within each age band and summarize them in a compact table, allowing me to compare whether the pattern of trait predictors is similar or different across emerging‑adult ages that map onto the target range for my planned chatbot study.


```{r - loading data}
library(tidyverse)
library(haven)

mTurk_raw <- read_sav("OnlinePsychologySurveyData.sav") %>% 
  as_tibble()
```

```{r - create dataframe to measure openness and big 5 traits, age, and gender scores}
big5 <- mTurk_raw %>% 
  transmute(
    openness           = OPENNESS,
    extraversion    = EXTRAVERSION,         
    agreeableness   = AGREEABLENESS,
    conscientiousness = CONSCIENTIOUSNESS,
    emotional_stability = EMOTIONALSTABILITY,
    age = Age,
    gender = Gender
  )
```

```{r - recoding gender}
big5 <- big5 %>% 
  mutate(
    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )

glimpse(big5)
```

```{r - loading psych}
library(psych)

cor_vars <- big5 %>%
  select(openness, extraversion, agreeableness, conscientiousness, emotional_stability)

cor_results <- psych::corr.test(cor_vars, use = "pairwise")
cor_results$r      # correlations
cor_results$p      # p-values

model_full <- lm(openness ~ extraversion + agreeableness +
                   conscientiousness + emotional_stability,
                 data = big5)

cor_obj <- psych::corr.test(cor_vars, use = "pairwise")
cor_mat <- cor_obj$r    

summary(model_full)
```
```{r - heat map}
cor_df <- cor_mat %>%
  as.data.frame() %>%
  tibble::rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r")

ggplot(cor_df, aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-1, 1), name = "r"
  ) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank()
  )
```

Interpretation: Using AI, I created a heatmap that helps visualize the correlations among openness and the other Big 5 traits, showing a clear visual that shows which of the traits are most closely aligned to openness. Darker red means a stronger positive relationship; paler pink means a weaker positive one. Openness correlates .36 with extraversion, .34 with agreeableness, .28 with conscientiousness, and .25 with emotional stability. This means people higher in openness in this sample also tend to be somewhat more extraverted, agreeable, conscientious, and emotionally stable, but the relationships are moderate, not huge.

```{r - single scatterplot with all smooth lines}
plot_data <- big5 %>%
  select(openness, extraversion, agreeableness,
         conscientiousness, emotional_stability) %>%
  mutate(across(everything(), scale)) %>%        # z-scores
  as.data.frame() %>%
  pivot_longer(
    cols = c(extraversion, agreeableness,
             conscientiousness, emotional_stability),
    names_to = "predictor",
    values_to = "x_value"
  )

ggplot(plot_data, aes(x = x_value, y = openness, color = predictor)) +
  geom_point(alpha = 0.15, size = 1) +
  geom_smooth(method = "lm", se = FALSE, size = 1.1) +
  scale_color_brewer(palette = "Dark2",
                     name = "Predictor",
                     labels = c("Agreeableness", "Conscientiousness",
                                "Emotional stability", "Extraversion")) +
  labs(x = "Standardized predictor score",
       y = "Standardized openness",
       title = "Big Five predictors of openness") +
  theme_minimal(base_size = 12)
```

Interpretation: Each line is showing the linear relationship between openness and a predictor, with each trait visualized through a different color. As shown in the figure, higher extraversion and emotional stability were most strongly associated with greater openness, with agreeableness and conscientiousness also positively related. This is because the steeper a line, the stronger the relationship. The upward slopes are saying that people who are higher on any of these traits also tend to be higher on openness in the sample. 


```{r - creating a single coefficient plot to show all predictors}
library(broom)

model_full <- lm(openness ~ extraversion + agreeableness +
                   agreeableness + conscientiousness +
                   emotional_stability,
                 data = big5)

coef_df <- tidy(model_full, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = recode(term,
                       extraversion      = "Extraversion",
                       agreeableness     = "Agreeableness",
                       conscientiousness = "Conscientiousness",
                       emotional_stability    = "Emotional stability"))

ggplot(coef_df,
       aes(x = estimate, y = term, xmin = conf.low, xmax = conf.high,
           color = term)) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray50") +
  geom_point(size = 2) +
  geom_errorbarh(height = 0.15, size = 0.9) +
  scale_color_brewer(palette = "Dark2", guide = "none") +
  labs(x = "Unstandardized regression coefficient",
       y = NULL,
       title = "Regression coefficients predicting openness") +
  theme_minimal(base_size = 12)
```

Interpretation: This plot shows how strongly each of the traits predicts openness when in the model all together. The lines estimate the 95% confidence interval. From this plot, it can be interpreted that all four traits have positive coefficients, so higher extraversion, emotional stability, conscientiousness, and agreeableness each predict higher openness when you adjust for the others. Since extraversion has the largest coefficient and its confidence interval is above 0, it would be the strongest unique predictor of openness in the model. Agreeableness and conscientiousness show significant positive associations and emotional stability show some modest contribution.




