---
title: "Portfolio 6 - Narcissiscm and Extraversion as Predictors"
---

### Goal

For my Research Methods 1 class, I wrote a paper looking at narcissism and extraversion as predictors of status seeking and group affiliation behaviors using the same mTurk data. In this portfolio piece, I want to make visualizations/figures of my main findings in order to demonstrate them in a visually appealing way that is not found in the original paper that I wrote. It will illustrate how narcissism and extraversion relate to group‑affiliation and status‑seeking motives in a sample of 300 adults, emphasizing that narcissism uniquely predicts status seeking whereas extraversion uniquely predicts group‑affiliation seeking.

### Products

The products will include a correlation heatmap showing associations among narcissism, extraversion, group‑affiliation, and status seeking, two scatterplots with smooth lines depicting the bivariate relationships of narcissism with group‑affiliation and status seeking and a pair of bar or coefficient plots summarizing the hierarchical regression results, highlighting that extraversion uniquely predicts group‑affiliation while narcissism uniquely predicts status seeking when both are entered together.



```{r - loading data}
library(tidyverse)
library(haven)

mTurk_raw <- read_sav("OnlinePsychologySurveyData.sav") %>% 
  as_tibble()
```
```{r - create dataframe to measure narcissism, extraversion, status seeking, and group affiliation}
narc_cor <- mTurk_raw %>% 
  transmute(
    narcissism           = NARCISSISM,
    extraversion    = EXTRAVERSION,         
    group_aff   = GROUP,
    status = STATUS,
    age = Age,
    gender = Gender
  )
```

```{r - recoding gender}
narc_cor <- narc_cor %>% 
  mutate(
    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )

glimpse(narc_cor)
```
```{r - cor matrix}
vars <- narc_cor %>%
  select(narcissism, extraversion, group_aff, status)

cor_obj <- psych::corr.test(vars, use = "pairwise")
cor_mat <- cor_obj$r

cor_df <- cor_mat %>%
  as.data.frame() %>%
  tibble::rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "r")
```

```{r - correlation heatmap}
ggplot(cor_df, aes(x = var1, y = var2, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3) +
  coord_fixed() +
  labs(x = NULL, y = NULL,
       title = "Correlations among narcissism, extraversion, and social motives") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```
Interpretation: From this heatmap, it can be interpreted that narcissism travels most strongly with status seeking, and extraversion travels most strongly with group‑affiliation. 

```{r - narcissism vs group aff}
ggplot(narc_cor, aes(x = narcissism, y = group_aff)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(x = "Narcissism",
       y = "Group-affiliation motive",
       title = "Narcissism and group-affiliation seeking") +
  theme_minimal(base_size = 12)
```

Interpretation: This scatterplot shows a modest positive association between narcissism and group‑affiliation motive. Participants higher in narcissism tended to report slightly stronger desires to belong to social groups, although the wide spread of points indicates substantial variability at each narcissism level. 


```{r - narcissism vs status seeking}
ggplot(narc_cor, aes(x = narcissism, y = status)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(x = "Narcissism",
       y = "Status-seeking motive",
       title = "Narcissism and status seeking") +
  theme_minimal(base_size = 12)
```

Interpretation: This scatterplot is showing a strong positive association between narcissism and status‑seeking motive where participants with higher narcissism scores tended to report substantially greater desire for social status and prestige, with relatively little overlap in predicted status‑seeking at the lowest versus highest levels of narcissism. It has a more upwards slope than group-affiliation seeking. 



```{r - hiearchial regression}
library(broom)

m1_group <- lm(group_aff ~ extraversion, data = narc_cor)
m2_group <- lm(group_aff ~ extraversion + narcissism, data = narc_cor)


m1_status <- lm(status ~ extraversion, data = narc_cor)
m2_status <- lm(status ~ extraversion + narcissism, data = narc_cor)


anova(m1_group, m2_group)
anova(m1_status, m2_status)
```
```{r - label cleaning, helped by AI}
coef_group <- tidy(m2_group, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    outcome = "Group affiliation",
    term = recode(term,
                  extraversion = "Extraversion",
                  narcissism   = "Narcissism")
  )

coef_status <- tidy(m2_status, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    outcome = "Status seeking",
    term = recode(term,
                  extraversion = "Extraversion",
                  narcissism   = "Narcissism")
  )

coef_all <- bind_rows(coef_group, coef_status)
```


```{r - faceted coefficient plots}
ggplot(coef_all,
       aes(x = estimate, y = term,
           xmin = conf.low, xmax = conf.high,
           color = term)) +
  geom_vline(xintercept = 0, linetype = 2, color = "gray") +
  geom_point(size = 2) +
  geom_errorbarh(height = 0.15, size = 0.9) +
  facet_wrap(~ outcome) +
  scale_color_brewer(palette = "Dark2", name = "Predictor") +
  labs(x = "Unstandardized regression coefficient",
       y = NULL,
       title = "Unique effects of narcissism and extraversion on social motives") +
  theme_minimal(base_size = 12)
```

Interpretation: Based on the visual from the group-affiliation side, extraversion acts as a unique predictor wanting group affiliation, but narcissism does not add meaningful prediction once extraversion is controlled. From the status-seeking side, narcissism uniquely predicts status seeking, whereas extraversion does not uniquely matter for status once narcissism is included. 




```{r - narcissism by gender}
ggplot(narc_cor, aes(x = gender, y = narcissism, fill = gender)) +
  geom_violin(alpha = 0.3, width = 0.9, trim = FALSE) +
  geom_boxplot(width = 0.3, outlier.alpha = 0.3) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = "Gender",
       y = "Narcissism (NPI score)",
       title = "Narcissism scores by gender") +
  theme_minimal(base_size = 12)
```

Interpretation: I have always heard that narcissism as a trait is found higher in males than females, so I wanted to create this figure out of curiosity. From it, we see the male median is higher, and more of the male density sits at moderate to high narcissism scores. The figure is showing that based on this sample, men tend to score higher on narcissism than women, though there is still overlap (some women are high, and some men are low).


