---
title: "Portfolio 6 - Narcissiscm and Extraversion as Predictors"
---

### Goal

For my Research Methods 1 class, I wrote a paper looking at narcissism and extraversion as predictors of status seeking and group affiliation behaviors using the same mTurk data. In this portfolio piece, I want to make visualizations/figures of my main findings in order to demonstrate them in a visually appealing way that is not found in the original paper that I wrote. It will illustrate how narcissism and extraversion relate to group‑affiliation and status‑seeking motives in a sample of 300 adults, emphasizing that narcissism uniquely predicts status seeking whereas extraversion uniquely predicts group‑affiliation seeking.

### Products

The products will include a correlation heatmap showing associations among narcissism, extraversion, group‑affiliation, and status seeking, two scatterplots with smooth lines depicting the bivariate relationships of narcissism with group‑affiliation and status seeking and a pair of bar or coefficient plots summarizing the hierarchical regression results, highlighting that extraversion uniquely predicts group‑affiliation while narcissism uniquely predicts status seeking when both are entered together.



```{r - loading data}
library(tidyverse)
library(haven)

mTurk_raw <- read_sav("OnlinePsychologySurveyData.sav") %>% 
  as_tibble()
```
```{r - create dataframe to measure narcissism, extraversion, status seeking, and group affiliation}
narc_cor <- mTurk_raw %>% 
  transmute(
    narcissism           = NARCISSISM,
    extraversion    = EXTRAVERSION,         
    group_aff   = GROUP,
    status = STATUS,
    age = Age,
    gender = Gender
  )
```

```{r - recoding gender}
narc_cor <- narc_cor %>% 
  mutate(
    gender = factor(gender,
                    levels = c(1, 2),
                    labels = c("Male", "Female"))
  )

glimpse(narc_cor)
```




